---
title: "Class 11: Protein Structure Prediction"
author: "Leah Kim A16973745"
format: pdf
toc: true
---

## pLDDT Score from Alignment File 

Here we read the results from AlphaFold and try to interpret all the models and quality score metrics.

```{r}
library(bio3d)

pth <- "hivprdimer_23119/"
pdb.files <- list.files(path = pth, full.names = TRUE, pattern = ".pdb")
```

Align and superpose all these models
```{r}
file.exists(pdb.files)
```

```{r}
pdbs <- pdbaln(pdb.files, fit = TRUE, exefile="msa")
```

```{r}
library(bio3dview)
#view.pdbs(pdbs)
```

```{r}
#model 5 is significantly lower than the otehr models, lower score
plot(pdbs$b[1,], typ="l", ylim = c(0,100), ylab = "PLDDT Score")
lines(pdbs$b[2,], typ="l", col="blue")
lines(pdbs$b[3,], typ="l", col="green")
lines(pdbs$b[4,], typ="l", col="orange")
lines(pdbs$b[5,], typ="l", col="red")
```

## Custom Analysis of Resulting Models
```{r}
rd <- rmsd(pdbs, fit=T)
range(rd)
```

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

```{r}
# Read a reference PDB structure
pdb <- read.pdb("1hsg")
```

```{r}
# Read a reference PDB structure
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

```{r}
core <- core.find(pdbs)
core.inds <- print(core, vol=0.5)
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

## Predicted Alignment Error
```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path=pth,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```

```{r}
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```

```{r}
# Per-residue pLDDT scores 
#  same as B-factor of PDB..
head(pae1$plddt) 
```

```{r}
pae1$max_pae
```

```{r}
pae5$max_pae
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
#plot.dmat(pae1$pae, 
          #xlab="Residue Position (i)",
          #ylab="Residue Position (j)",
          #grid.col = "black",
          #zlim=c(0,30)
```

## Score residue Conservation from Alignment file

AlphaFold returns it's large alignment file used for analusis. Here we read this file and score conservation per position.

```{r}
aln_file <- list.files(path=pth, pattern=".a3m$", full.names = TRUE)
aln_file
```
```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

How many sequences are in this alignment

```{r}
dim(aln$ali)
```

We can score residue conservation in the alignment with the conserv() function.

```{r}
sim <- conserv(aln)

plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

Export file for viewing in Mol
```{r}
m1.pdb <- read.pdb(pdb.files[1])
occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)
write.pdb(m1.pdb, o=occ, file="m1_conserv.pdb")
```

