---
title: "Class 12"
author: Leah Kim A16973745
format: pdf
toc: true
---

# Class 12 Homework 

## Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. Hint: The read.table(), summary() and boxplot() functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein. Note that you can find the medium value from saving the output of the boxplot() function to an R object and examining this object. There is also the medium() and summary() function that you can use to check your understanding. 

Reading the dataset
```{r}
data <- read.table("rs8067378_ENSG00000172057.6.txt", header = TRUE, sep = "", stringsAsFactors = TRUE)
names(data)
```

Sample Size for each genotype: 
```{r}
table(data$geno)
```

## Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need to be as polished as this one.

Boxplot for each genotype:
```{r}
bp <- boxplot(exp ~ geno, data = data, 
              main = "Expression by Genotype",  
              xlab = "Genotype", 
              ylab = "Expression")
bp$stats[3, ]
```

Median expression of each genotype: 
```{r}
bp$stats[3, ]
```

As shown by the boxplot, the A/A genotype has the highet median expression of around 30, the A/G has the middle expression of around 25, and G/G has the lowest median expression of around 20. Though there is a bit of overlap, there is a consistent decline in expression the more G alleles the genotype has.

This suggests that the SNPS does affect the expression of ORMDL_3 with the presense of the G allele reducing expression. Individuals with G/G genotype consistently show lower expression, while those with A/A show the highest. 

Summary of data:
```{r}
summary(data)
```
