---
title: "Class18: Investigating pertussis Resurgence Mini-Project"
author: "Leah Kim A16973745"
format: pdf
toc: true
---

## Background

Pertussis (aka whooping cough) is a common lung infection caused by the bacteria *B. Pertussis*. 

The CDC tracs cases of Pertussis in the US:
https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html 

##Examininc cases of Pertussis by year

We can use the **datapasta** package to scrape case numbers from the CDC website
```{r, include = FALSE}
cdc <- data.frame(Year = c(1922L,1923L,1924L,1925L,
                           1926L,1927L,1928L,1929L,1930L,1931L,
                           1932L,1933L,1934L,1935L,1936L,
                           1937L,1938L,1939L,1940L,1941L,1942L,
                           1943L,1944L,1945L,1946L,1947L,
                           1948L,1949L,1950L,1951L,1952L,
                           1953L,1954L,1955L,1956L,1957L,1958L,
                           1959L,1960L,1961L,1962L,1963L,
                           1964L,1965L,1966L,1967L,1968L,1969L,
                           1970L,1971L,1972L,1973L,1974L,
                           1975L,1976L,1977L,1978L,1979L,1980L,
                           1981L,1982L,1983L,1984L,1985L,
                           1986L,1987L,1988L,1989L,1990L,
                           1991L,1992L,1993L,1994L,1995L,1996L,
                           1997L,1998L,1999L,2000L,2001L,
                           2002L,2003L,2004L,2005L,2006L,2007L,
                           2008L,2009L,2010L,2011L,2012L,
                           2013L,2014L,2015L,2016L,2017L,2018L,
                           2019L,2020L,2021L,2022L,2023L,2024L),
                  Cases = c(107473,164191,165418,152003,
                            202210,181411,161799,197371,
                            166914,172559,215343,179135,265269,
                            180518,147237,214652,227319,103188,
                            183866,222202,191383,191890,109873,
                            133792,109860,156517,74715,69479,
                            120718,68687,45030,37129,60886,
                            62786,31732,28295,32148,40005,
                            14809,11468,17749,17135,13005,6799,
                            7717,9718,4810,3285,4249,3036,
                            3287,1759,2402,1738,1010,2177,2063,
                            1623,1730,1248,1895,2463,2276,
                            3589,4195,2823,3450,4157,4570,
                            2719,4083,6586,4617,5137,7796,6564,
                            7405,7298,7867,7580,9771,11647,
                            25827,25616,15632,10454,13278,
                            16858,27550,18719,48277,28639,32971,
                            20762,17972,18975,15609,18617,
                            6124,2116,3044,7063,35493)
)
```

>. Q Make a plot of pertussis cases per year using ggplot

```{r}
library(ggplot2)

cases <- ggplot(cdc) +
  aes(x = Year, y = Cases) + 
  geom_point() +
  geom_line()

cases
```

> Q2. Add some key time points in our history of interaction with Pertussis. These include wP rollout (the first vaccine) in 1946 and the switch to aP in 1996, and the COVID lockdown in 2020.

We can use `geom_vline()` for this.

```{r}
cases +
  geom_vline(xintercept = 1946, col = "blue") + 
  geom_vline(xintercept = 1996, col = "red") +
  geom_vline(xintercept = 2020, col = "green")
```

Mounting evidence suggests that the newer **aP** vaccine is less effective over the loong term than the older **wP** vaccine that it replaced.In other words, vaccine proection wanes more rapidly with aP or with wP.

## Enter the CMI-PB Project

The goal of the CMI-PB (Computational Models of Immunity - Pertussis Boost) project is to investigate how the immune responds differently to aP vs wP vaccinated individuals and be able to predict this at an early stage.

CMI-PB makes all their collected data freely available and they store it in a databse composed of different tables. Here, we will access a few of these.

We can use the **jsonlite** package to read this data.

```{r}
library(jsonlite)

subjects <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)

head(subjects)
```

> Q. How many subjects are there?

```{r}
nrow(subjects)
```

> Q. How many aP and wP subjects are there?

```{r}
table(subjects$infancy_vac)
```

> Q. How many males nd females are there in the dataset?

```{r}
table(subjects$biological_sex)
```


> Q. How about gender and race?

```{r}
table(subjects$race, subjects$biological_sex)
```

> Q. Is this representative of the US population?

No. This is more the demographic of UCSD students (since this data was gathered from UCSD students).


Let' read another database table from CMI-PB.

```{r}
specimens <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)

ab_data <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)
```

Wee peek at these:
```{r}
head(specimens)
```

```{r}
head(ab_data)
```

We want to join these tabels to get all our information togehter. For this, we will use the **dplyr** package and the `inner_join()` function.
```{r}
library(dplyr)

meta <- inner_join(subjects, specimens)
head(meta)
```

One more join to get ab_data and mta all together.

```{r}
abdata <- inner_join(ab_data, meta)
head(abdata)
```

```{r}
dim(abdata)
```

> Q. How many antibody isotypes are there in the dataset?

```{r}
table(abdata$isotype)
```

> Q. How many different antigens are measured in the dataset?

```{r}
table(abdata$antigen)
```
> Q. Make a boxplot of antigen levels across the whole dataset.

```{r}
ggplot(abdata) +
  aes(x = MFI, y = antigen) + 
  geom_boxplot()
```

Measles is flat as it is meant to be a kind of control. Meanwhile, antigens such as TT and ACT are present since the vaccine contains them.

> Q. Are there obvious differences between aP and dP values?

```{r}
ggplot(abdata) +
  aes(x = MFI, y = antigen, col = infancy_vac) + 
  geom_boxplot()
```

There are some differences but nothing that significant.

## Focus on IgG level

IgG is the most abundant antibody in blood. With four sub-classes (IgG1 to IgG4) crucial for long term-immunity and responding to bacterial and viral infections.

```{r}
igg <- abdata |> filter(isotype == "IgG")
head(igg)
```

Same boxplot of anitgens as before.

```{r}
ggplot(igg) +
  aes(x = MFI_normalised, y = antigen, col = infancy_vac) + 
  geom_boxplot() + facet_wrap(~visit)
```

Focus in further on just one of these antigens - Let's pick PT (Pertussis toxin, one of the mian toxins of the bacteria).

```{r}
table(igg$dataset)
```

```{r}
pt_igg <- abdata |> filter(isotype == "IgG", antigen == "PT", dataset == "2021_dataset")
dim(pt_igg)
```

```{r}
ggplot(pt_igg) + aes(x = actual_day_relative_to_boost, y = MFI_normalised, col = infancy_vac, group = subject_id) + 
  geom_point() +
  geom_line() +
  theme_bw() +
  geom_vline(xintercept = 0) + 
  geom_vline(xintercept = 14, col = "red")
```

On day 14, there is a visible peak in the MFI, and the paths show that generally, wP individuals have a higher MFI normalized value than aP.

